---
title: "17SEP24_data_wrangling_i"
author: "Stella Koo"
date: "2024-09-15"
output: github_document
---

### Two kinds of paths:
* Absolute: a file or folder's "full address" on the computer
    * If directory moved, or project shared, absolute path to files will change and code won’t work
* Relative: directions to a file or folder from the current working directory
    * Portable, if directory moved or shared, the relative path will stay the same and code will work

### Shorthand Notation for Pathnames
* `~` Home Directory
* `.` Current working directory
* `..` One directory up from current working directory
* `../..` Two directories up from current working directory
* `gitignore` file: specifies intentionally untracked files that Git should ignore.

### Importing data tables
Use function from `readr` (included in tidyverse) to import data.

* `janitor::clean_names()`: standardizes column names by converting them to a consistent, machine-friendly format
* `package::function` allows using a function from a package without loading the whole library (e.g. `dplyr::filter`)
* `conflicted` package: helps resolve ambiguity by forcing to explicitly choose which function to use when there's a conflict

```{r}
library(tidyverse)

litters_df = read_csv(file = "./data_import_examples/FAS_litters.csv") # relative path
names(litters_df) 
# e.g Litter Number

litters_df = janitor::clean_names(litters_df)
names(litters_df)
# with janitor::clean_names() function -> litter_number
```
```{r}
pups_df = read_csv(file = "~/Desktop/Columbia Biostatistics MS/Courses/P8105/Lecture Notes/17SEP24_data_wrangling_i/data_import_examples/FAS_pups.csv") # absolute path
```

### Looking at data
* `view`: opens a spreadsheet-like view of data frame (does not work well inside RMD files)
* `str`: displays structure of R object (e.g data types of each column for data frames)
* `head`:returns the first few rows of data frame
* `tail`: returns the last few rows of data frame
* `skimr::skim()` function: part of the skimr package, which provides a more comprehensive summary of data frames 

```{r}
tail(litters_df, 5)
```
```{r}
skimr::skim(litters_df)
```

### Arguments to `read*`
* `col_names`: default is TRUE (first row of the dataset is automatically used as the column names). FALSE generic column names like `X1`, `X2`, ..., are assigned
* `na`: specify which values in data should be treated as missing values. (e.g. na = "N/A")
* `skip`: skips a specified number of rows before reading the data

```{r}
litters_df_2 = 
    read_csv(file = "./data_import_examples/FAS_litters.csv",
    skip = 10, col_names = FALSE)

head(litters_df_2)
```
```{r}
litters_df_3 = 
    read_csv(
        file = "./data_import_examples/FAS_litters.csv",
        na = c(".", "NA", ""))

head(litters_df_3)
```

### Parsing columns
`read_* functions` (e.g. `read_csv`) will automatically guess the data type of columns. However, if you want to give explicit column specifications use `cols` functions:

```{r}
litters_df_4= 
    read_csv(file = "./data_import_examples/FAS_litters.csv",
        na = c(".", "NA", ""),
    col_types = cols(
      Group = col_factor(),
      `Litter Number` = col_character(),
      `GD0 weight` = col_double(),
      `GD18 weight` = col_double(),
      `GD of Birth` = col_integer(),
      `Pups born alive` = col_integer(),
      `Pups dead @ birth` = col_integer(),
      `Pups survive` = col_integer()
    )
  )

tail(litters_df_4)
```

```{r}
pups_df_2 = read_csv(file = "./data_import_examples/FAS_pups.csv",
                     na = c(".", "NA"),
                     col_types = "fddddd" # factor, double, double, ...
)

head(pups_df_2)
```

### Other file formats
For Excel files, use `readxl` package that has `read_excel` function.
* `sheet`option: useful when there are multiple sheets in the Excel file
* `range` option: want to read in a specific data rectangle

```{r}
library(readxl)

mlb11_df = read_excel("./data_import_examples/mlb11.xlsx", n_max = 20)

head(mlb11_df, 5)
```

For SAS, Stata, SPSS files, use `haven` package.
```{r}
library(haven)

pulse_df = read_sas("./data_import_examples/public_pulse_data.sas7bdat")

head(pulse_df, 5)
```

### Comparisons with Base R
`read.csv`, `read.table`, etc are base R functions (different from `read_csv` that come from `readr` package, outputs a tibble, and does NOT convert character cols to factors).
* tend to be slower
* outputs a data frame
* automatically convert characters to factors (less control over data types)

```{r}
pups_base = read.csv("./data_import_examples/FAS_pups.csv")
pups_readr = read_csv("./data_import_examples/FAS_pups.csv")

head(pups_base, 5)
head(pups_readr, 5)

pups_base$S
pups_readr$S
```

## Exporting data
Use `write_*` function in `readr` package.
(e.g. write_csv(data_frame, "output.csv"))
