---
title: "24SEP24_tidy_data"
author: "Stella Koo"
date: "2024-09-23"
output: github_document
---

# Functions in `tidyr` package (in `tidyverse`)

```{r message = FALSE}
library(tidyverse)
options(tibble.print_min = 5)
```

Use `haven` package to load Pulse Data with SAS file type:
```{r}
pulse_df = 
  haven::read_sas("./data_import_examples/public_pulse_data.sas7bdat") |>
  janitor::clean_names()

pulse_df
```

## `pivot_longer()`
Use `pivot_longer()` function that makes datasets longer by increasing number of rows and decreasing the number of columns:
```{r}
pulse_tidy_df = 
  pivot_longer(
    pulse_df,
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi"
  )

pulse_tidy_df
```

* `names_prefix`: removes common prefix from the column names that are pivoting into the "long" format.
* `values_to`: specifies the name of the new column that will store the values from the pivoting columns.

The following shows a tidier dataset by changing `bl` to `00m`, converting `visit` to a factor variable, and rounding `bdi` to integers. 
```{r}
pulse_df = 
  haven::read_sas("./data_import_examples/public_pulse_data.sas7bdat") |>
  janitor::clean_names() |>
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") |>
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit),
    bdi = round(bdi))

print(pulse_df, n = 12)
```

#### Learning Assessment
```{r}
litter_tidy_df = 
  read_csv("./data_import_examples/FAS_litters.csv") |>
  janitor::clean_names() |>
  pivot_longer(
    gd0_weight:gd18_weight,
    names_to = "gd",
    values_to = "weight") |>
  mutate(gd = recode(gd, gd0_weight = 0, gd18_weight = 18)) |>
  select(litter_number, gd, weight)

head(litter_tidy_df)
```

## `pivot_wider()`
`pivot_wider()` transforms data from a long to a wide format. Takes values from a column and spreads them across multiple columns.

```{r}
analysis_result = 
  tibble(
    group = c("treatment", "treatment", "placebo", "placebo"),
    time = c("pre", "post", "pre", "post"),
    mean = c(4, 8, 3.5, 4)
  )

analysis_result
```

```{r}
pivot_wider(
  analysis_result, 
  names_from = "time", 
  values_from = "mean")
```

## Binding rows
Non-tidiness often stems from relevant data spread across multiple tables. These tables are basically the same and can be stacked to produce a tidy dataset.

```{r}
fellowship_ring = 
  readxl::read_excel("./data_import_examples/LotR_Words.xlsx", range = "B3:D6") |>
  mutate(movie = "fellowship_ring")

two_towers = 
  readxl::read_excel("./data_import_examples/LotR_Words.xlsx", range = "F3:H6") |>
  mutate(movie = "two_towers")

return_king = 
  readxl::read_excel("./data_import_examples/LotR_Words.xlsx", range = "J3:L6") |>
  mutate(movie = "return_king")
```

```{r}
lotr_tidy =
  bind_rows(fellowship_ring, two_towers, return_king) |>
  janitor::clean_names() |>
  pivot_longer(
    female:male,
    names_to = "gender", 
    values_to = "words") |>
  mutate(race = str_to_lower(race)) |> 
  select(movie, everything()) 

lotr_tidy
```

## Joining Datasets
Data can be spread across multiple related tables, in which case it is necessary to combine or join them prior to analysis.

* Inner: keeps data that appear in both x and y
* Left: keeps data that appear in x
* Right: keeps data that appear in y
* Full: keeps data that appear in either x or y

```{r message = FALSE}
pup_df = 
  read_csv(
    "./data_import_examples/FAS_pups.csv",
    na = c("NA", "", ".")) |>
  janitor::clean_names() |>
  mutate(
    sex = 
      case_match(
        sex, 
        1 ~ "male", 
        2 ~ "female"),
    sex = as.factor(sex)) 

litter_df = 
  read_csv(
    "./data_import_examples/FAS_litters.csv",
    na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  separate(group, into = c("dose", "day_of_tx"), sep = 3) |>
  relocate(litter_number) |>
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    dose = str_to_lower(dose))
```

* `separate(group, into = c("dose", "day_of_tx"), sep = 3)`: splits the group column into two new columns: dose and day_of_tx. The split occurs after the third character.
* `relocate(litter_number)`: moves the litter_number column to the front of the dataframe.

```{r}
fas_df = 
  left_join(pup_df, litter_df, by = "litter_number")

fas_df
```

#### Learning Assessment

* `str_replace_all(id, "student_", "")`: removes "student_" from the rows in `id` column. (from `stringr` package in `tidyverse`)

```{r}
surv_od_df = 
  read_csv("./survey_results/surv_os.csv",
    na = c("NA", ".", "")) |>
  janitor::clean_names() |> 
  rename(id = what_is_your_uni, os = what_operating_system_do_you_use) |>
  mutate(id = str_replace_all(id, "student_", ""))

surv_program_df = read_csv("./survey_results/surv_program_git.csv",
    na = c("NA", ".", "")) |>
  janitor::clean_names() |> 
  rename(
    id = what_is_your_uni, 
    prog = what_is_your_degree_program,
    git_exp = which_most_accurately_describes_your_experience_with_git) |>
    mutate(id = str_replace_all(id, "student_", ""))

surv_df = inner_join(surv_od_df, surv_program_df)
print(surv_df, 10)
```

The `inner_join()` gives warnings about “many-to-many” relationships. These indicate that the matching variable is not unique. Row 7 of x matches multiple rows in y:

```{r}
surv_od_df |> slice(7)
surv_program_df |> filter(id == "15")
```

* `slice(7)`: from the `dplyr` package, extracts row 7 from `surv_od_df` dataframe.
* `filter(id == "15")`: filters rows with `id == 15`

